Structure Learning for $\VAR(p)$ Models

### 2019-09-30 presentations/student_dynamical_systems_waterloo_2019/Abstract-Template.tex
abstract for dynamic systems workshop

\begin{document}
\hspace{-0.70cm}\rule{17.5cm}{0.01 in} \\
\vspace{-0.4cm}
\begin{flushleft}
\Large \textbf{\noindent
Insert the Title of Your Talk Here}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\\
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vspace{0.5cm}
\normalsize
 %%%%%%             NAMES OF AUTHORS         %%%%%%
 %%%%%% NAME OF SPEAKER SHOULD BE UNDERLINED %%%%%%%%%
\normalsize{
 \underline{Ryan J. Kinnear}$^1$, Ravi R. Mazumdar$^1$
} \\
\vspace{5mm}
\textit{\footnotesize
 %%%%%% AFFILIATION OF AUTHORS %%%%%%
$^1$ Department of Electrical and Computer Engineering, University of Waterloo, Waterloo, Ontario N2L 3G1, Canada\\
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
}
\end{flushleft}

We study Granger causality in the context of wide-sense stationary
time series, where our focus is on the topological aspects of the
underlying causality graph.  We establish sufficient conditions (in
particular, we develop the notion of a ``strongly causal'' graph
topology) under which the true causality graph can be recovered via
pairwise causality testing alone, and provide examples from the gene
regulatory network literature suggesting that our concept of a
strongly causal graph may be applicable to this field.  We implement
and detail finite-sample heuristics derived from our theory, and
establish through simulation the efficiency gains (both statistical
and computational) which can be obtained (in comparison to LASSO-type
algorithms) when structural assumptions are met.

In this paper we study the notion of Granger causality
\cite{granger1969investigating} \cite{Granger1980329} as a means of
uncovering an underlying causal structure in multivariate time series.
Though the underlying causality graph cannot be observed directly,
it's presence is inferred as a latent structure among observed
time series data.  This notion is leveraged in a variety of
applications e.g. in Neuroscience as a means of recovering
interactions amongst brain regions \cite{bressler2011wiener},
\cite{anna_paper2008}, \cite{david2008identifying}; in the study of
the dependence and connectedness of financial institutions
\cite{NBERw16223}; gene expression networks \cite{Fujita2007},
\cite{methods_for_inferring_gene_regulatory_networks_from_time_series_expression_data},
\cite{grouped_graphical_granger_modelling_for_gene_expression_regulatory_networks_discovery},
\cite{discovering_graphical_Granger_causality_using_the_truncating_lasso_penalty};
and power system design \cite{Misyrlis2016450}, \cite{yuan2014root}.

The principal contributions of this paper are as follows: firstly, in
section \ref{sec:theory} we study \textit{pairwise} Granger causality
relations, providing novel theorems connecting the structure of the
causality graph to the pairwise ``causality flow'' in the system, as
well as an interpretation in terms of the graph topology of the
sparsity pattern of matrices arising in the Wold decomposition,
generalizing in some sense the notion of ``feedback-free'' processes
studied by \cite{caines1975feedback} in close connection with
Granger causality.  We establish sufficient conditions (sections
\ref{sec:strongly_causal_graphs}, \ref{sec:persistent_systems}) under
which a fully conditional Granger causality graph can be recovered
from pairwise tests alone (sec \ref{sec:pairwise_algorithm}).  We
report a summary of simulation results in \ref{sec:simulation}, with
additional results reported in the supplementary material Section
\ref{apx:simulation}.  Our simulation results establish that there is
significant potential for improvement over existing methods, and that
the graph-topological aspects of time series analysis are relevant for
both theory and practice.  Concluding remarks on further open problems
and extensions are provided in Section \ref{sec:conclusion}.  The
proofs of each proposition and theorem are also relegated to the
supplementary material, simple corollaries have proofs included in the
main text.

In a general Granger causality graph, there are natural
\textit{necessary} conditions for a node $j$ to pairwise cause node
$i$, namely that $j$ is an ancestor to $i$ or that there is a third
confounding node $k$.  On the other hand, we will see simple examples
for why \textit{sufficient} conditions are more nuanced: if $j$ causes
$i$ conditionally on all other nodes in the graph, it need not be the
case that $j$ causes $i$ when the two are observed in isolation.  We
examine a special \textit{graph topological} assumption we term
``strongly causal'' wherein a natural converse \textit{does} hold,
satisfying an intuitive notion of ``causality flow'' through the
graph.  We highlight examples from the literature on gene networks
suggesting that assumptions of strongly causal graph structure may not
be out of place in some application areas.

As a consequence of the causality graph being strongly causal, we show
that it is possible to recover the full conditional causality graph
through pairwise testing alone -- that is, it isn't necessary to
perform any conditional causality tests.

% $\bullet$ Abstracts may contain equations: Equations are to be centered horizontally, with equation numbers aligned with the right margin. See equation (\ref{eq1})
% \begin{equation}
% e^{i\pi}+1=0.\label{eq1}
% \end{equation}

% $\bullet$ Abstracts must contain at least two  references. See   \cite{ASY,STE}.

% $\bullet$ Abstracts may contain figures. For example, Figure \ref{fig1}.

\begin{figure}[H]
\centering
\framebox[6cm]{\rule{0cm}{0.5cm} Abstracts may contain figures}
%\includegraphics[width =0.48\linewidth]{image} 
\caption{Figure one caption.}
\label{fig1}
\end{figure}

 Please send your source file  and .pdf file to Alyssa Novelia (anovelia@uwaterloo.ca) or Isam Al-Darabsah (ialdarabsah@uwaterloo.ca)

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% BIBLIOGRAPHY
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{thebibliography}{00}
\addcontentsline{toc}{chapter}{References}
\setlength{\itemsep}{-1mm}
\small
\bibitem{ASY} K.T. Alligood, T.D. Sauer and J. A. Yorke. \textit{Chaos: An Introduction to Dynamical Systems}, New York, Springer-Verlag (1997).

\bibitem{STE} L. Schimansky-Geier, A.V. Tolstopjatenko and W. Ebelin \textit{Noise induced transitions due to external additive noise}, Phys. Lett. A, \textbf{108}, 7, pp. 329-332 (1985).
\end{thebibliography}
\normalsize
\end{document}

>>>>

% \usepackage{graphicx}
% \usepackage{caption}
\usepackage{subcaption}
\newtheorem{theorem}{Theorem}

\def\gcg{\mathcal{G}}  % Granger-causality graph

\begin{document}
\hspace{-0.70cm}\rule{17.5cm}{0.01 in} \\
\vspace{-0.4cm}
\begin{flushleft}
  \Large \textbf{\noindent Graph Topological Aspects of Granger
    Causality in Stationary Linear Dynamical Systems}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\\
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vspace{0.5cm}
\normalsize
 %%%%%%             NAMES OF AUTHORS         %%%%%%
 %%%%%% NAME OF SPEAKER SHOULD BE UNDERLINED %%%%%%%%%
\normalsize{
 \underline{Ryan J. Kinnear}$^1$, Ravi R. Mazumdar$^1$
} \\
\vspace{5mm}
\textit{\footnotesize
 %%%%%% AFFILIATION OF AUTHORS %%%%%%
$^1$ Department of Electrical and Computer Engineering, University of Waterloo, Waterloo, Ontario N2L 3G1, Canada\\
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
}
\end{flushleft}

The concept of Granger causality, a notion introduced in 1969
\cite{granger1969investigating} for the analysis of macroeconomic data
(e.g. ``does lowering the interest rate cause an increase in
employment?''), attempts to capture the intuition that a cause must
proceed it's effect, and has spurred both theoretical and applied
research.  The concept has been applied in diverse settings, for
example: in finance, researchers have attempted to quantify the
interdependence of financial institutions \cite{NBERw16223} and in the
biological sciences, the concept is used to map out the promoting and
regulating effects of certain genes upon others
\cite{methods_for_inferring_gene_regulatory_networks_from_time_series_expression_data}.

A substantial portion of the work on this topic is devoted to
generalizing the methods of Granger to nonlinear systems.  For
example, linear Granger causality is a special case of the information
theoretic notion of transfer entropy \cite{barnett2009granger}.
However, like many attempts at quantifying causation, transfer entropy
suffers from some difficulties of interpretation
\cite{transfer_entropy_criticism}.  Although the criticisms raised by
\cite{transfer_entropy_criticism} do not arise in the case of linear
Granger causality, other pathologies are known.  Our work sheds light
on some of the challenges that remain in the linear case, and in
particular focuses on topological particularities of the underlying
network of interactions.

In general, we are presented with a multitude of jointly wide sense
stationary time series $x_1(t), \ldots, x_n(t)$ and seek to construct
a (Granger-)causal graph $\gcg$ where there is an edge
$(j, i) \in \gcg$ whenever $x_j$ Granger causes $x_i$ after
conditioning on each $k \not\in \{i, j\}$.  Depending on the
interpretation or on the application, the graph $\gcg$ captures the
flows of ``information'' or ``energy'' throughout a network of
interacting processes.  However, it may be the case that we observe
$x_j$ and $x_i$ alone -- in such cases, simple examples establish that
even if $x_j$ Granger causes $x_i$ conditioned on the rest of the
observations, it need not be that $x_j$ Granger causes $x_i$ pairwise
(i.e. without conditioning).

Pathologies of pairwise causation clash with our intuition that causal
relations in $\gcg$ should be transitive, that is, that if there is
some ``true'' (conditional) causation path
$j \rightarrow a_0 \rightarrow \cdots \rightarrow a_r \rightarrow i$
in $\gcg$ then it should hold that $x_j$ pairwise causes $x_i$.  While
this does not hold in general, our work analyzes a particular type of
graph topology, which we deem a ``strongly causal'' graph, wherein
natural intuitions about causal flow \textit{do} hold.  Moreover, we
prove that (with modest additional technical assumptions) that in such
a graph the complete conditional causation graph $\gcg$ can be
recovered by testing only pairwise interactions of nodes.  That is,
the complete causal structure of $\gcg$ can be recovered through
simple one way unconditional causality tests.

The flavour of this work may be conveyed by stating one of our main
conclusions, where the ideas behind $T_{ij}(z)$ are subtle and not
expanded upon in this brief summary.

\begin{theorem}[Pairwise Recovery]
  \label{thm:scg_recovery}
  If the Granger causality graph $\gcg$ for the process $x(t)$ is
  a strongly causal DAG and $T_{ij}(z)$ is a two-sided filter for every
  $i, j \in [n]$, then $\gcg$ can be recovered from pairwise causality
  tests alone.
\end{theorem}

% \begin{figure}[h]
%   \centering
%   \begin{subfigure}[b]{0.45\textwidth}
%     \includegraphics[width=\linewidth]{../../figures/example_algorithm.pdf}
%   \end{subfigure}
%   \begin{subfigure}[b]{0.45\textwidth}
%     \includegraphics[width=\linewidth]{../../figures/example_algorithm2.pdf}
%   \end{subfigure}
% \end{figure}

\begin{figure}[h]
  \centering
  \caption{Recovering Large Networks via Pairwise Causality Tests}
  \label{fig:large_n_figure}
  \includegraphics[width=0.7\linewidth]{../../figures/new_increasing_n_simulation.pdf}
\end{figure}

Our concept of a strongly causal graph topology initially arose
artificially as a technical condition to understand the intuition of
causality flow in $\gcg$ and to establish Theorem
\ref{thm:scg_recovery}.  However, examples from the gene regulatory
network literature suggest that the dynamics of some natural
interacting systems may have an underlying interaction graph which is
in fact strongly causal, or at least nearly so.  Simulation evidence
also suggests that our algorithm for graph recovery via pairwise
testing can outperform the widely employed LASSO (and variations
thereon) for systems which satisfy (or nearly satisfy) the strongly
causal condition; in the process we make the incidental observation
that, all else equal, the LASSO exhibits superior performance on the
strongly causal topology than on a general directed acyclic graph,
suggesting that topological aspects of the underlying causality graph
may be of interest as an aspect unique to time series data in
comparison to ordinary regression.  Finally, the results of Figure
\ref{fig:large_n_figure} show that when topological assumptions are
met, it is possible to accurately identify extremely large networks
($n > 1000$, implying more than $10^6$ possible edges), and that the
underlying graph topology is an important determination of estimation
accuracy in addition to the simple size and sparsity of the network.

 % Please send your source file  and .pdf file to Alyssa Novelia (anovelia@uwaterloo.ca) or Isam Al-Darabsah (ialdarabsah@uwaterloo.ca)

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% BIBLIOGRAPHY
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{thebibliography}{00}
\addcontentsline{toc}{chapter}{References}
\setlength{\itemsep}{-1mm}
\small
\bibitem{granger1969investigating} Granger,
  C. W. J. (1969). \textit{Investigating causal relations by
    econometric models and cross-spectral methods.} Econometrica 37
  (3), 424–438.

\bibitem{NBERw16223} Billio, M., M. Getmansky, A. W. Lo, and
  L. Pelizzon (2010, July). \textit{Econometric measures of systemic
    risk in the finance and insurance sectors.}  Working Paper 16223,
  National Bureau of Economic Research.

\bibitem{methods_for_inferring_gene_regulatory_networks_from_time_series_expression_data}
  Nguyen, P. (2019). \textit{Methods for Inferring Gene Regulatory
    Networks from Time Series Expression Data.}
  Ph. D. thesis, Northwestern University

\bibitem{barnett2009granger} Barnett, L., A. B. Barrett, and
  A. K. Seth (2009). \textit{Granger causality and transfer entropy
    are equivalent for gaussian variables.} Physical review letters
  103(23), 238701.

\bibitem{transfer_entropy_criticism} Ryan G James, Nix Barnett, and
  James P Crutchfield. \textit{Information flows? A critique of
    transfer entropies}. In: Physical review letters 116.23 (2016),
  p. 238701
\end{thebibliography}
\normalsize
\end{document}
